Controlling the dewetting morphologies of thin liquid films by switchable substrates

### 2022-11-25 Reply_RefereeA.tex

\section*{Reply to Referee A}
We thank the Referee for his/her report on our manuscript. 
We are glad to read that he/she finds it of potential interest and in principle 
suitable of publication, if a thorough revision is performed. 
We substantially improved and revised the paper for the resubmission and hope that we were able to clarify the obscure points, following all the Referee's remarks.
Hereafter we provide answers to the various questions/comments raised. All changes in the revised version are reported in red colour.

\begin{itemize}

\item[ \textbf{\underline{Comment 1.}}]
{ 
I am unable to follow most of the arguments made with respect to the disjoining pressure. 
For me, one needs to integrate the disjoining pressure to arrive at an effective interface potential, and 
I do not really know how to go from the effective interface potential to the surface tension. 
The equation (3) therefore merits much more discussion.
}

\item[ \textbf{{Answer}}]
{
We did not give, in fact, any particular argument with respect to the disjoining pressure, since 
we assumed that it is a very well established concept employed in any theoretical and computational study of thin liquid film hydrodynamics in the lubrication approximation framework.
We also assumed that an equation like Eq.(3) (now Eq.(2)), which is approximately twenty-five years old \cite{SCHWARTZ1998173,Mitlin94}, did not deserve a lengthy discussion.
However, our assumptions were probably erroneous and, therefore, we give, here, some more explanation. 
Let us, first, make a step back and underline a fundamental difference between the physics of {\it one free interface} and that of a {\it film}:
the latter is a layer of liquid confined between {\it two} interfaces. When the two interfaces are brought close enough to each other (the layer is thin), they interact.
The disjoining pressure stems from this interaction. Thermodynamically, it represents minus the derivative of the free energy difference, per unit surface, between 
the film and the bulk of the same liquid, or, equivalently, the pressure difference inside and outside the film 
\cite{Deryaguin1940,DeryaguinChuraev1978}.
In the case of wetting, there can be three types of interfaces: liquid/air, liquid/solid (the substrate) and, 
if the substrate is only partially wetted, solid/air, thus giving rise to three-phase contact lines. 
The usually called {\it surface tension} is the energy cost of forming a liquid/air interface of unit area and its only knowledge is, therefore, not enough in our case. 
The Young's condition \cite{Young1805,DeGennes1985} connects the surface tension with the analogous energies per unit
area associated with the solid/liquid ($\gamma_{\text{sl}}$) and solid/gas ($\gamma_{\text{sg}}$) interface as
\begin{equation}
\cos \theta = \frac{\gamma_{\text{sg}}-\gamma_{\text{sl}}}{\gamma},
\end{equation}
where $\theta$ is the contact angle.
So, there is no need (and no way) to go from the disjoining pressure (or the interface potential) to the surface tension, since the two quantities carry complementary physical information,
and both are required to determine the dynamics of dewetting.
Of course, on the other hand, the disjoining pressure depends also on the surface tension, which appears, in fact, in its expression (3). 
We think, though, that part of the confusion might be due to the fact that we do not define the symbol of the surface tension at its first occurrence, but only later. 
We have now fixed this and we apologize for it.
}

@@

\item[ \textbf{\underline{Comment 2.}}] 
{
The same holds for the time scale. 
The viscous time scale is well defined, but necessitates a length scale in addition to the 
capillary velocity. It is unclear why the authors choose the length scale that they use in the paper.
Subsequently, they replace the surface tension by a complicated partial derivative of the 
disjoining pressure; here the same remark as above holds.
}

\item[ \textbf{{Answer}}]
{
We apologize if we have not been thorough in the discussion of the relevant scales, which indeed deserves some clarification. 
Thin film hydrodynamics is an intrinsically multiscale phenomenon, whereby there are at least two characteristic length scales: the mean film thickness (let us call it $h_0$)
and an horizontal length scale $\ell$ (which in our case can be, for example, the pattern wavelength, $\lambda$), where, typically $h_0 \ll \ell$. We chose to use the mean 
film thickness as a scale, because it is a relevant length that is common to all the simulations.
In the same spirit, we chose as a reference time $t_0(\theta)$, which is not exactly the capillary time (though being related to it). 
$t_0$, as we explain in the text, is, in fact, the characteristic time of growth of the most unstable mode in the spinodal dewetting on a homogeneous
substrate of contact angle $\theta$. In spite of the substrate being patterned in our case, it is possible to define an effective $t_0$ in terms 
of, e.g., the mean contact angle on the pattern, i.e. $t_0(\theta_0)$. 
We deem the so defined characteristic time as an appropriate scale,
since it does not depend on the pattern wavelength and speed. Actually, one can see that it captures well the time scales of the dewetting process
from figure 2, where it is shown that the rupture times are of the order of fraction to units of $t_0$.
We recognize, on the other hand that in this perspective the choice of $v_0 = \lambda/t_0$ as a reference velocity looks a bit awkward.
A more natural candidate would be, indeed, $v_0=\lambda_s/t_0$, where $\lambda_s$ is the spinodal 
characteristic length corresponding to $\theta_0$. We have therefore decided to use the latter in the revised version of the manuscript.
We have also rephrased the corresponding paragraph in the {\it Method} section as follows: \\

\textcolor{red}{Lengths and time scales will be expressed, respectively, in units of the mean film height, $h_0$ (which is constant in time, due to mass conservation), and of $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}$, the inverse growth rate of the most unstable mode, whose wavenumber is $q_0$, of a spinodally dewetting film. On a uniform substrate, with constant contact angle $\theta^{(u)}$, the wavenumber reads 
$(q^{(u)}_0)^2 = h_{\ast}^{-2}(1-\cos \theta^{(u)})f^{\prime}(h_0/h_{\ast})$ \cite{Mecke_2005,PhysRevE.100.023108}. In our patterned case, we define 
$q_0^2=h_{\ast}^{-2}(1-\cos\theta_0)f^{\prime}(h_0/h_{\ast})$.
Correspondingly, we choose as a velocity scale $v_0 = \lambda_s/t_0$, where $\lambda_s = 2\pi/q_0$.}

@@

}

\item[ \textbf{\underline{Comment 3.}}]
{ 
Incidentally, for a 'normal' fluid such as water, the capillary velocity is $\sim 70\mbox{m}/\mbox{s}$, 
so that very, very rapidly the fluid inertia becomes important. 
This is indeed what one usually observes for the formation or coalescence of drops, so that at the very least this should be discussed and the Reynolds number calculated.
}

\item[ \textbf{{Answer}}]
{
Such a high value corresponds to the ratio of (surface tension)/(dynamic viscosity) for water, i.e. $v = \gamma/\eta \approx 70 \mbox{m}/\mbox{s}$.
This expression, though, is only representative of the capillary velocity, again, for {\it free surface flows}, as, e.g., in the coalescence of 
two droplets in the bulk of another fluid. 
In the case wetting/dewetting, 
the phenomenology is complicated by the liquid/substrate (and vapour/substrate) interaction. 
In such a situation the relevant velocity scale is set by the characteristic speed of contact line 
motion which can be estimated from a viscocapillary balance as 
$\sim \varepsilon^3 \gamma/\eta$, where $\varepsilon = h_0/L \ll 1$ is the film thickness parameter 
\cite{RevModPhys.69.931,doi:10.1146/annurev-fluid-011212-140734}. 
The typical velocities in dewetting problems are, then, much smaller than the ratio $\gamma/\eta$.
Moreover, in many actual applications, as well as in the majority of experiments, the 
working fluids are in general much more viscous than water (e.g., polymeric liquids), thus making
the characteristic velocity even smaller.
For the dewetting of a millimetric film of micrometric thickness reatracting into a single droplet, for instance, experiments have measured that 
the typical velocity is $\sim 1 \mbox{mm}/\mbox{s}$, which gives a Reynolds number $Re \approx 10^{-3}$ \cite{Edwards2016}. 
In our problem, there are two relevant characteristic velocities: the retraction speed, 
$U_{\Theta}$, and 
the pattern speed, $v_{\theta}$. When their ratio $\Gamma = v_{\theta}/U_{\Theta} < 1$, the retraction speed dominates and fixes the Reynolds number which can be, then, evaluated as 
$Re = U_{\Theta}h_0/\nu$ \cite{RevModPhys.69.931}, whereas for $\Gamma > 1$, $Re = v_{\theta} h_0/\nu$. 
In any case, the Reynolds number stays always below $\approx 0.2$.
We have added the following comment at the end of the {\it Method} section:\\

@@

\item[ \textbf{\underline{Comment 4.}}]
{ 
In addition, there is a lot of old literature that relates the dewetting to the disjoining pressure, and shows that the nucleation of a hole in a wetting film is much more difficult than the nucleation of such a wetting film; there are experiments on this by Rutledge and Taborek, theory by Schick and Taborek and Bausch, Blossey and Indekeu.
}

\item[ \textbf{{Answer}}]
{
We are well aware of the literature quoted by the Referee, but these are works 
about {\it prewetting}, namely the phenomenon of {\it nucleation} (condensation)
of films/droplets on solid surfaces confining a critical liquid-vapour mixture
{\it as a function of temperature}. So, we are a bit puzzled about what such literature may have to do with the {\it dewetting} of a film on a chemically patterned substrate. Here, in fact, we are modelling a thin liquid film which is 
thermodaynamically stable against evaporation (i.e. one can consider that 
we are at a - fixed - temperature below the critical one).
}

\item[ \textbf{\underline{Comment 5.}}]
{ 
The explanations also lack a discussion on time, length, and energy scales. 
I have no clue what the values mean ``the numerical values, in LB units, are set to $\gamma= 0.01$ and $\mu = 1/6$''?. 
}

\item[ \textbf{{Answer}}]
{
Length and time scales, as discussed in Comment 2 and as mentioned also in the original version of the manuscript, 
are made dimensionless with the mean film height and the characteristic time $t_0$. 
These two quantities 
have a clear physical meaning and can, therefore, be used to compare with 
experimental setups. In particular, the expression of $t_0$, through its dependence 
on $q_0$ (the wavenumber of the most unstable mode) contains the details 
of the disjoining pressure, such as the Hamaker constant $A_H$. For instance, let us consider the problem studied in \cite{becker2003complex}, the dewetting of a polystyrene film of thickness $h_0 \approx 4 \, \text{nm}$ deposited on a silicon 
dioxide substrate. The Hamaker constant for this system is 
$A_H  = 2.2 \times 10^{-20} \, \text{J}$ and the surface tension is 
$\gamma = 0.03 \, \text{N}/\text{m}$, such that, 
from equation (3) of \cite{becker2003complex} and 
from the definition of $q_0^2 = \frac{1}{2\gamma} \Pi^{\prime}(h_0)$, 
we get (neglecting the short range part of the interface potential)
$$
q_0^2 \approx \frac{A_H}{4\pi \gamma h_0^4} \approx 2.5 \times 10^{-4} \, \text{nm}^{-2}.
$$
The characteristic time of the growth of the instability, then, reads \cite{PhysRevLett.99.114503} (the dynamic viscosity of polystyrene is 
$\mu = 1.2 \times 10^4 \, \text{Pa} \cdot \text{s}$)
$$
t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4} \approx 300 \, \text{s}
$$
and it can be seen that indeed film rupture occurs within time 
scales of the order of $t_0$ ($\approx 2 t_0$) \cite{becker2003complex}.
We have reported these comments on comparison with actual experimental 
values in the new version of the paper as follows:\\
\\
\textcolor{red}{To give a flavour of how to compare with actual experimental systems, let us notice
first that the expression of $t_0$ depends on the disjoining pressure (through its dependence 
 $q_0 = \sqrt{\frac{1}{2\gamma} \Pi^{\prime}(h_0)}$) and contains, therefore, energetic details, namely 
 the Hamaker constant $A_H$. Consider, for instance, the problem studied in \cite{becker2003complex}: the dewetting of a polystyrene film of thickness $h_0 \approx 4 \, \text{nm}$ deposited on a silicon 
dioxide substrate, where $A_H  = 2.2 \times 10^{-20} \, \text{J}$ and 
$\gamma = 0.03 \, \text{N}/\text{m}$. Neglecting the short range part of the interface potential
(such that $\Pi(h)= - \frac{A_H}{6\pi h^3}$)~\cite{Mecke_2005,becker2003complex}, one gets 
$q_0^2 \approx \frac{A_H}{4\pi \gamma h_0^4} \approx 2.5 \times 10^{-4} \, \text{nm}^{-2}$.
The characteristic time of the growth of the instability for these experiments, then, reads \cite{PhysRevLett.99.114503} (the dynamic viscosity of polystyrene is 
$\mu = 1.2 \times 10^4 \, \text{Pa} \cdot \text{s}$) $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4} \approx 300 \, \text{s}$ and it can be seen that indeed film rupture occurs within time 
scales of the order of $t_0$ ($\approx 2 t_0$) \cite{becker2003complex}, as in our 
simulation for the largest pattern wavelength, $\lambda = L$ (see also our previous work
with homogeneous substrates~\cite{PhysRevE.104.034801}).} 
\\

Concerning the values of viscosity and surface tension in lattice Boltzmann units, 
we acknowledge that giving them in the middle of the discussion on characteristic length and time might  
have caused confusion. Nevertheless we deem important to provide them, for the sake of readers interested 
in numerical details of the method as well as for the results reproducibility, therefore we have moved them 
to a later paragraph, where we added: \\
\\
\textcolor{red}{The numerical values in lattice Boltzmann units of the 
mean film thickness, dynamics viscosity and surface tension are, respectively, $h_0=1$, $\mu=1/6$
and $\gamma=0.01$.}
}

### 2022-11-25 main_AS_final.tex
Please work

\begin{equation}\label{eq:mobility}
  M_{\delta}(h) = \frac{2h^3 + 6\,\delta\, h^2 + 3\delta^2h}{6\mu},

@@

The film pressure $p(\mathbf{x},t)$ consists of the sum of the Laplace and disjoining pressures, that is $p(\mathbf{x},t) = -\gamma \nabla^2 h - \Pi$, 

@@

\textcolor{red}{where $f(\xi)=\xi^{-3} - \xi^{-2}$} and $h_{\ast}$ is the height at which the disjoining pressure vanishes and sets the precursor layer thickness.

@@

@@

where $q_{\theta} = 2\pi/\lambda$, i.e.~the pattern evolves in time as a plane wave. 

@@

$q_0^2=h_{\ast}^{-2}(1-\cos\Theta)f^{\prime}(h_0/h_{\ast})$, where 
$\Theta = \max_{\mathbf{x}} \theta(\mathbf{x}) = \theta_0 + \delta \theta$.
A velocity scale then also naturally arises, namely $v_0 = h_0/t_0$.}

@@

    \caption{Stationary film thickness field ($t>20t_0$) showing the formation of droplets. The color map indicates the contact angle pattern 

@@

@@

$L/h_0 = 512 $.
To regularize the contact line divergence~\cite{huh1971hydrodynamic} we use a precursor layer thickness $h_{\ast}/h_0=0.07$ and a slip length $\delta/h_0 = 1$ (see Eqs.~(\ref{eq:mobility})-(\ref{eq:disjoinpressure})). 
The liquid film is initialized with a height field slightly perturbed around the mean value $h_0$, i.e.
\begin{equation}\label{eq:hinitial}
    h(\mathbf{x},0) = h_0 \left[1 + 0.1 \left(\sin\left(\frac{2\pi x}{L}\right)\sin\left(\frac{2\pi y}{L}\right)\right)\right].
Various wavelengths, in the range \textcolor{red}{$\lambda/h_0 \in [50, 512]$}, and velocities, $v_{\theta}/v_0 \in [0, 5 \times 10^3]$, are considered for the wettability pattern, Eq.~(\ref{eq:sinetheta}). 
\textcolor{red}{A comment on the limits of the wavelengths range is noteworthy at this point. The wavelengths are obviously limited from above by the size of the computational domain. On the other extreme, small values are prevented by
the limited capability of accomodating droplets of too large height, $h_d$, on too
small patches, since the local contact angle would possibly exceed the largest substrate contact angle, such that the patterning 
itself would loose meaning. In fact, if the patches were much smaller than 
the intrinsic size the droplets would have when formed on a homogeneous substrate
(fixed by the spinodal characteristic length, $\lambda_s$), the confinement by the 
pattern would not be effective~\cite{KarguptaPRL2001}
Besides, the very same lubrication approximation 
requires small contact angles. 
We keep, therefore, $\lambda \geq \lambda_s \approx 70 h_0$.}

@@

In Fig.~\ref{fig:model_rt} we report the rupture times, as a function of the wavelength, for stationary ($v_{\theta}=0$) and time-dependent ($v_{\theta}=10 v_0$) patterns. 

@@

In the stationary case we observe that $\tau_r$ grows linearly with $\lambda$, for short wavelengths, and quadratically for longer $\lambda$.
Also for $v_{\theta}=10$ the rupture times tend to approach the $\lambda^2$ scaling at large $\lambda$, whereas they tend to saturate to a constant value for short wavelengths.

@@

    \includegraphics[width=0.45\textwidth]{Figures/low_lambda_rupture.pdf}

@@

    \caption{Rupture times, $\tau_r$, as a function of the pattern wavelength $\lambda$, for $v_{\theta}=0$ (\textcolor{jlblue}{$\bullet$}) and $v_{\theta}=10 v_0$ 

@@

Interestingly, this phenomenology is qualitatively supported also by the time-dependent case, $v_{\theta} = 10v_0$ (reported in Fig.~\ref{fig:model_rt} as orange stars), \textcolor{red}{but for a smaller prefactor in the linear regime, due to the fact that the pattern wave velocity is large enough to dominate over $U_{\Theta}$ and sets the retraction speed, such that the characteristic time becomes $\tau_R \sim \lambda/v_{\theta}$.}

@@

    \includegraphics[width=0.45\textwidth]{Figures/static_pattern_test.pdf}

@@

    with $v_{\theta}= 0$ and $\lambda=L$ (blue circles), $\lambda=L/2$ (orange squares) and $\lambda=L/3$ (green stars).

@@

As a further consistency check, we follow the evolution of the quantity
\begin{equation}\label{eq:deltah}
  \Delta h(t) = \max_{\mathbf{x}}\{h(\mathbf{x},t)\}-\min_{\mathbf{x}}\{h(\mathbf{x},t)\}.
In the steady state, since the droplets are essentially monodisperse, this observable represents a measure of the mean droplet height $h_d$. Assuming that the droplet shape is a spherical cap, it can be estimated as $h_d =\left(\frac{3V_d(1-\cos(\theta_d))}{\pi(2+\cos(\theta_d))}\right)^{1/3}$, where $V_d = \frac{h_0 \lambda^2}{2}$ is the droplet volume and $\theta_d$ is the local contact angle (i.e. the value at the contact line). This, in turn, depends on $h_d$, through Eq.~(\ref{eq:sinetheta}), thus making the one above an implicit equation to be solved numerically for $h_d = h_d(\lambda)$. 
$\Delta h(t)$ is reported in Fig.~\ref{fig:clusters_v0_sine} for three different wavelengths, $\lambda = \{L/3,L/2,L \}$, together with the theoretically expected $h_d(\lambda)$ (depicted with lines), showing excellent agreement.\\

@@

    \includegraphics[width=0.45\textwidth]{Figures/lam_2_inset.pdf}
    \caption{Time evolution of the second order Minkowski structure metric, $q_2(t)$ (see text for details), for different $\Gamma$ values (see Eq.~(\ref{eq:vel_ratio})), on a substrate with pattern wavelength $\lambda=L/2$.

@@

For $v_{\theta} = 0.1 v_0$ we still observe the formation of droplets, similarly to the stationary case ($v_{\theta} = 0$). However, these are transported with the contact angle minima, reproducing a somehow similar behaviour recently described in a numerical study of a droplet on a moving wettability step~\cite{GrawitterStark1}.
If the pattern speed is further increased, for $v_{\theta} = v_0$ we observe the development of rivulet-like structures, aligned with $\mathbf{v}_{\theta}$. 

@@

In order to better characterize the various morphologies we apply the theory of Minkowski's functionals. 
In particular, we employ the second order Minkowski structure metric, $q_2$~\cite{doi:10.1063/1.4774084, Schaller2020}, which can be computed from a Voronoi tessellation of the set of discrete points $(x_i, y_i)$ on the 2D lattice, such that the height field lies above a certain threshold\footnote{The expression is: $q_2 = \frac{1}{N}\sum_j \frac{1}{P_j}|\sum_k L^{(j)}_k e^{2i\phi^{(j)}_k}|$, where the inner sum runs over the edges of length $L^{(j)}_k$, of the $j$-th Voronoi cell, whose perimeter is $P_j$, and $\phi^{(j)}_k$ is the polar angle of the normal to the $k$-th edge. 
The outer sum represents an ensemble average over the $N$ points in the set.}. The $q_2$ metric quantifies the degree of anisotropy of the dewetting morphology, so it takes relatively large values if the structures formed display a preferential direction. 
Measuring $q_2$ then enables us to clearly distinguish between the formation of droplets and rivulets: much larger $q_2$ values are attained for the latter type of structure, as we can see in Fig.~\ref{fig:msm_q2}.
%Our simulations with $\lambda = L/2$ show a clear difference between $v_{\theta} \leq 0.1v_0$ and $v_{\theta}>0.1v_0$  in $q_2$.
We observe, on the other hand, that such rivulets are metastable and eventually break up into droplets, as indicated by the collapse of $q_2$ at later times. 
Notice, though, that the $q_2$ signal for any $v_{\theta} >0$ always stays above the one for the static case, suggesting that even the smallest pattern velocity introduces a sizeable deformation of the spherical cap shape.
\begin{figure}
    \centering
    % \includegraphics[width=0.45\textwidth]{Figures/Lig_stab_inset_delta_h.pdf}
    \includegraphics[width=0.45\textwidth]{Figures/Gamma_riv_lifetime.pdf}
    \caption{MAIN PANEL: Rivulets life-times, $\tau_{\text{riv}}$, for various $\Gamma$ values.
    The dashed line is plotted as a guide to the eye to highlight the logarithmic dependence, in agreement with 
    the theoretical prediction, Eq.~(\ref{eq:rivlt}).
    INSET: Height fluctuations, $\Delta h(t)$, vs time, computed along the rivulet axis, for three different $\Gamma$ values.
    }
    \label{fig:stab_ligs_lam2}
\end{figure}
The breakup is the result of a varicose mode of the rivulet~\cite{doi:10.1063/1.3211248, PhysRevE.77.061605}, whose wavelength is $\approx\lambda$, such that only $N_{\infty}/2$ droplets are counted after breakup. 
These droplets show a peculiar dynamics, characterized by a periodic sequence of spreading and retraction, driven by the pattern, that we dub "pumping state" (see movie file ligament\_formation\_and\_breakup.mp4 in the Supp. Mat.).

>>>>

$M_{\delta}(h) = \frac{2h^3 + 6\,\delta\, h^2 + 3\delta^2h}{6\mu}$

@@

The film pressure $p(\mathbf{x},t)$ consists of the sum of the Laplace and disjoining pressures, that is $p(\mathbf{x},t) = -\gamma \nabla^2 h - \Pi$~\footnote{\textcolor{red}{The Laplace pressure is the product of the surface tension 
and the local curvature, therefore its full expression reads $p_L = -\gamma \nabla^2 h/\sqrt{1+|\nabla h|^2}$; 
since we are in the lubrication regime, though, whereby $|\nabla h| \ll 1$, the denominator can be safely neglected 
and the pressure reduces to the form involving only the Laplacian of the thickness 
field~\cite{Benet2014,alizadeh_pahlavan_cueto-felgueroso_hosoi_mckinley_juanes_2018}.}}, 

@@

\textcolor{red}{where $f(\xi)=\xi^{-3} - \xi^{-2}$} and $h_{\ast}$ is the height at which the disjoining pressure vanishes and sets the precursor layer thickness~\footnote{\textcolor{red}{Expressions of 
the general form $\Pi(h) = \gamma(1-\cos \theta)(n-1)(m-1)/(n-m)((h_{\ast}/h)^n - (h_{\ast}/h)^m)/h_{\ast}$
are frequently found in the literature, with various combination of exponents $(n,m)$ ($n>m$), depending
on the type of interactions among fluid molecules and between fluid and substrate~\cite{SCHWARTZ1998173,Mitlin,Teletzke}. A common choice, for instance, is the pair $(9,3)$, which 
can be derived from molecular interactions of Lennard-Jones type. Here, the use of the $(3,2)$ pair is motivated mainly by the computational advantage of keeping the characteristic dewetting time 
shorter. We have run also a single case ($\lambda=256 h_0$ and $\Gamma=1.5$) with $(9,3)$ and observed that 
the dynamics remains essentially the same, but for a shift in the time scale (see also~\cite{PhysRevE.104.034801})}}.

@@

\textcolor{red}{We decorate the substrate with a checkerboard pattern, a common choice that generalizes the broken homogeneity of the stripes to two 
directions~\cite{KarguptaLangmuir2003,Jalali2018,Nagayama2020,Das2020}.}

@@

where $q_{\theta} = 2\pi/\lambda$, i.e.~the pattern evolves in time as a plane wave~\footnote{\textcolor{red}{A laboratory experiment for the here mentioned approach consists of a light switchable substrate with a thin film cast on it~\cite{becker2003complex, IchimuraEtAl_Science2000}. 
The external stimulus, our moving pattern, can be realized by using a digital multimirror device (DMD)~\cite{doi:10.1021/jp301092y,doi:10.1126/science.aax8760}.}}. 

@@

  $q_0^2=h_{\ast}^{-2}(1-\cos\theta_0)f^{\prime}(h_0/h_{\ast})$. Correspondingly, we choose as a velocity scale $v_0 = \lambda_s/t_0$, where
$\lambda_s = 2\pi/q_0$.}

@@

    \caption{Stationary film thickness field ($t>4t_0$) showing the formation of droplets. The color map indicates the contact angle pattern 

@@

%%%%
\textcolor{red}{To give a flavour of how to compare with actual experimental systems, let us notice
first that the expression of $t_0$ depends on the disjoining pressure (through its dependence 
 $q_0 = \sqrt{\frac{1}{2\gamma} \Pi^{\prime}(h_0)}$) and contains, therefore, energetic details, namely 
 the Hamaker constant $A_H$. Consider, for instance, the problem studied in \cite{becker2003complex}: the dewetting of a polystyrene film of thickness $h_0 \approx 4 \, \text{nm}$ deposited on a silicon 
dioxide substrate, where $A_H  = 2.2 \times 10^{-20} \, \text{J}$ and 
$\gamma = 0.03 \, \text{N}/\text{m}$. Neglecting the short range part of the interface potential
(such that $\Pi(h)= - \frac{A_H}{6\pi h^3}$)~\cite{Mecke_2005,becker2003complex}, one gets 
$q_0^2 \approx \frac{A_H}{4\pi \gamma h_0^4} \approx 2.5 \times 10^{-4} \, \text{nm}^{-2}$.
The characteristic time of the growth of the instability for these experiments, then, reads \cite{PhysRevLett.99.114503} (the dynamic viscosity of polystyrene is 
$\mu = 1.2 \times 10^4 \, \text{Pa} \cdot \text{s}$) $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4} \approx 300 \, \text{s}$ and it can be seen that indeed film rupture occurs within time 
scales of the order of $t_0$ ($\approx 2 t_0$) \cite{becker2003complex}, as in our 
simulation for the largest pattern wavelength, $\lambda = L$ (see also our previous work
with homogeneous substrates~\cite{PhysRevE.104.034801}).}
%%%%

@@

$L/h_0 = 512 $. \textcolor{red}{The numerical values in lattice Boltzmann units of the 
mean film thickness, dynamics viscosity and surface tension are, respectively, $h_0=1$, $\mu=1/6$
and $\gamma=0.01$.}
\textcolor{red}{To regularize the contact line divergence~\cite{huh1971hydrodynamic} we use a precursor layer thickness $h_{\ast}/h_0=0.07$ and a slip length $\delta/h_0 = 1$, 
a value which lies
within the weak/intermediate slip regime~\cite{peschka2019signatures,fetzer2007quantifying, munch2005lubrication} (we have tried also a value four times smaller, without observing 
any major qualitative change in the results).} 
The liquid film is initialized with a height field slightly perturbed around the mean value $h_0$, i.e. $h(\mathbf{x},0) = h_0 \left[1 + 0.1 \left(\sin\left(\frac{2\pi x}{L}\right)\sin\left(\frac{2\pi y}{L}\right)\right)\right]$.
Various wavelengths, in the range \textcolor{red}{$\lambda/h_0 \in [64, 512]$}, and velocities,
$v_{\theta}/v_0 \in [0, 200]$, are considered for the wettability pattern, Eq.~(\ref{eq:sinetheta}). 
\textcolor{red}{A few comments on these ranges are noteworthy at this point. The wavelengths are obviously limited from above by the size of the computational domain. On the other extreme, too short wavelengths are prevented by
the energetic cost of accomodating droplets on too
small patches, since the local contact angle 
(that increases with decreasing $\lambda$~\cite{SuppMat}) would possibly exceed the largest substrate contact angle, such that the patterning 
itself would loose meaning. 
It has been, in fact, proven, both theoretically and experimentally, that 
for the confinement to be effective, the pattern characteristic size should 
not be much smaller than the spinodal length scale on the less wettable substrate
\cite{KarguptaJCP2002,KarguptaLangmuir2000,Nisato1999,Karim1998}.
We keep, therefore, $\lambda \stackrel{>}{\sim} \lambda_s \approx 70 h_0$.
The pattern speeds, on the other hand, are limited by the intrinsic LB bound of the lattice speed of sound
 \cite{succi}.}\\

@@

In Fig.~\ref{fig:model_rt} we report the rupture times, as a function of the wavelength, for stationary ($v_{\theta}=0$) and time-dependent ($v_{\theta}=20 v_0$) patterns. 

@@

We observe that $\tau_r$ grows linearly with $\lambda$, for short wavelengths, and quadratically for longer $\lambda$.

@@

    \includegraphics[width=0.45\textwidth]{Rupture_times_Gamma_theta_20.pdf}

@@

    \caption{Rupture times, $\tau_r$, as a function of the pattern wavelength $\lambda$, for $v_{\theta}=0$ (\textcolor{jlblue}{$\bullet$}) and $v_{\theta}=20 v_0$ 

@@

%Interestingly, this phenomenology is qualitatively supported %also by the time-dependent case, $v_{\theta} = 20 v_0$ (reported %in Fig.~\ref{fig:model_rt} as orange stars), \textcolor{red}{but %for a smaller prefactor in the linear regime, due to the fact %that the pattern wave velocity is large enough to dominate over %$U_{\Theta}$ and sets the retraction speed, such that the %characteristic time becomes $\tau_R \sim \lambda/v_{\theta}$.}

@@

    \includegraphics[width=0.45\textwidth]{Static_new_theta20t0.pdf}

@@

    with $v_{\theta}= 0$ and $\lambda= 512 h_0$ (blue circles), $\lambda=256 h_0$ (orange squares) and $\lambda=170 h_0$ (green stars).

@@

\textcolor{red}{In the main panel of the same figure, we see that the height fluctuations $\Delta h(t) = \max_{\mathbf{x}}\{h(\mathbf{x},t)\}-\min_{\mathbf{x}}\{h(\mathbf{x},t)\}$
grow in time until film rupture and then settle to a constant value, which represents a measure of the mean droplet height $h_d$ (since
droplets are essentially monodisperse), decreasing with the pattern wavelength (as expectable, due to the decrease of the droplet volume,
$V_d = \frac{h_0 \lambda^2}{2}$).}\\
%In the steady state, since the droplets are essentially monodisperse, this observable . Assuming that the droplet shape is a spherical cap,
%it can be estimated as $h_d =\left(\frac{3V_d(1-\cos(\theta_d))}{\pi(2+\cos(\theta_d))}\right)^{1/3}$, where  is the droplet volume and
%$\theta_d$ is the local contact angle (i.e. the value at the contact line). This, in turn, depends on $h_d$,
%through Eq.~(\ref{eq:sinetheta}), thus making the one above an implicit equation to be solved numerically for $h_d = h_d(\lambda)$. 
%$\Delta h(t)$ is reported in Fig.~\ref{fig:clusters_v0_sine} for three different wavelengths, 
%$\lambda/h_0 = \{170,256,512 \}$, together with the theoretically expected $h_d(\lambda)$ (depicted with lines),
%showing excellent agreement.\\

@@

    \includegraphics[width=0.45\textwidth]{lam_2_inset_theta20.pdf}
    \caption{Time evolution of the second order Minkowski structure metric, $q_2(t)$ (see text for details), for different $\Gamma$ values, on a substrate with pattern wavelength $\lambda=256 h_0$.

@@

For $v_{\theta} = 0.2 v_0$ we still observe the formation of droplets, similarly to the stationary case ($v_{\theta} = 0$). However, these are transported with the contact angle minima, reproducing a somehow similar behaviour recently described in a numerical study of a droplet on a moving wettability step~\cite{GrawitterStark1}.
If the pattern speed is further increased, for $v_{\theta} = 2 v_0$ we observe the development of rivulet-like structures, aligned with $\mathbf{v}_{\theta}$. 

@@

In order to better characterize the various morphologies we apply the theory of Minkowski's functionals. 
In particular, we employ the second order Minkowski structure metric, $q_2$~\cite{doi:10.1063/1.4774084, Schaller2020}, which can be computed from a Voronoi tessellation of the set of discrete points $(x_i, y_i)$ on the 2D lattice, such that the height field lies above a certain threshold\footnote{The expression is: $q_2 = \frac{1}{N}\sum_j \frac{1}{P_j}|\sum_k L^{(j)}_k e^{2i\phi^{(j)}_k}|$, where the inner sum runs over the edges of length $L^{(j)}_k$, of the $j$-th Voronoi cell, whose perimeter is $P_j$, and $\phi^{(j)}_k$ is the polar angle of the normal to the $k$-th edge. 
The outer sum represents an ensemble average over the $N$ points in the set.}. The $q_2$ metric quantifies the degree of anisotropy of the dewetting morphology, so it takes relatively large values if the structures formed display a preferential direction. 
Measuring $q_2$ then enables us to clearly distinguish between the formation of droplets and rivulets: much larger $q_2$ values are attained for the latter type of structure, as we can see in Fig.~\ref{fig:msm_q2}.
%Our simulations with $\lambda = L/2$ show a clear difference between $v_{\theta} \leq 0.1v_0$ and $v_{\theta}>0.1v_0$  in $q_2$.
We observe, on the other hand, that such rivulets are metastable and eventually break up into droplets, as indicated by the collapse of $q_2$ at later times. 
Notice, though, that the $q_2$ signal for any $v_{\theta} >0$ always stays above the one for the static case, suggesting that even the smallest pattern velocity introduces a sizeable deformation of the spherical cap shape.
\begin{figure}
    \centering
    % \includegraphics[width=0.45\textwidth]{Figures/Lig_stab_inset_delta_h.pdf}
    \includegraphics[width=0.45\textwidth]{lig_lifetime_Gamma_theta20.pdf}
    \caption{MAIN PANEL: Rivulets life-times, $\tau_{\text{riv}}$, for various $\Gamma$ values.
    The dashed line is plotted as a guide to the eye to highlight the logarithmic dependence, in agreement with 
    the theoretical prediction, Eq.~(\ref{eq:rivlt}).
    INSET: Height fluctuations, $\Delta h(t)$, vs time, computed along the rivulet axis, for three different $\Gamma$ values.
    }
    \label{fig:stab_ligs_lam2}
\end{figure}
The breakup is the result of a varicose mode of the rivulet~\cite{doi:10.1063/1.3211248, PhysRevE.77.061605}, whose wavelength is $\approx\lambda$, such that only $N_{\infty}/2$ droplets are counted after breakup. 
These droplets show a peculiar dynamics, characterized by a periodic sequence of spreading and retraction, driven by the pattern, that we dub "pumping state"~\footnote{see movie file ligament\_formation\_and\_breakup.mp4 in~\cite{SuppMat}.}.
